---
title: "Code Chunk Options"
author: "Dr. Austin Brown"
output: html_document
---

Let's take a look at some different options we have with respect to code chunks! Let's say we want to generate the penguin plot. We can create it in an `R Markdown` document quite easily:

```{r}
library(tidyverse)
library(palmerpenguins)
palmerpenguins::penguins |>
  ggplot(aes(bill_length_mm,bill_depth_mm)) +
  geom_point()
```

Okay here we notice several things. First, when we attach the `tidyverse` package, we get the messages output to the knitted document as well. Additionally, we also get the warning messages about the points being omitted. Let's keep that in mind.  

What if we want to simply include code, but don't want it to run? We can change the default setting in our code chunk from `eval=TRUE` to `eval=FALSE`.

```{r,eval=FALSE}
## This won't execute ##
palmerpenguins::penguins |>
  ggplot(aes(bill_length_mm,bill_depth_mm)) +
  geom_point()
```

If we wanted to repress the code and repress the output, which might be ideal if you were creating functions or attaching libraries, we can change the default chunk option `include=TRUE` to `include=FALSE`. The code will still execute, but nothing will appear in your document.

```{r,include=FALSE}
## This will execute but won't render ##
p <- palmerpenguins::penguins |>
  ggplot(aes(bill_length_mm,bill_depth_mm)) +
  geom_point()
```
```{r}
p
```

Quite commonly, when we're giving professional presentations, we might not want to include the code, but for sure want the output. Here, we can change the default chunk option `echo=TRUE` to `echo=FALSE`.

```{r,echo=FALSE}
## This will execute but the code won't render ##
palmerpenguins::penguins |>
  ggplot(aes(bill_length_mm,bill_depth_mm)) +
  geom_point()
```

The other two main chunk code options I utilize all the time are `message = FALSE` and `warning = FALSE`. What these do is repress, you guessed it, the warning messages and other messages the appear in the console. So remember how when we attached the `tidyverse` package that we got that output the typically prints to the console? You likely won't want that in a professional presentation!

```{r,echo=TRUE,warning=FALSE,message=FALSE}
## No Tidyverse Messages or Warning Messages will Appear! ##
library(tidyverse)
palmerpenguins::penguins |>
  ggplot(aes(bill_length_mm,bill_depth_mm)) +
  geom_point()
```

There are lots of other chunk options out there for controlling various aspects of your output (including things like figure captions, size of the output, etc.). For a full list, check out <https://rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf?_ga=2.234246238.1518898826.1600885588-865726554.1588115827>.  

Now, depending on your application, you may always use the same code chunk options in a single `R Markdown` file. So going through each code chunk and specifying the options you want can increase the amount of time spent on the  analysis/write-up itself. To get around this, we can specify global options at the onset of our markdown file. By default, if you open a new `R Markdown` file in RStudio, you'll get a global code chunk code in the initial document:

```{r setup,echo=TRUE,eval=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

So for me, because I always use `message=FALSE,warning=FALSE`, I just specify them up here in this code so I don't have to repeatedly do it for each code chunk (pretty handy!).